Nextstrain: real-time tracking of virus evolution

### 2017-10-09 whitepaper.tex
minor improvements

Nextstrain consists of three components:
``Fauna'' is a collection of python scripts to maintain a database representative of all available sequences and related metadata, sourced from public repositories such as NCBI, gisaid and ViPR, as well as github repositories and other sources of genomic data. ``Augur'' is the bioinformatics pipeline, performing subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait reconstruction across the tree, including inference of the most likely transmission events.
This leverages the maximum likelihood phylodynamic analysis TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n=000$) in 0 hours.
These scripts separate generic core functionality from a light virus specific layer such that they are easily co-opted to different pathogens.
Analysed data are then immediately available to interrogate through the visualisation software ``Auspice'', available at \href{www.nextstrain.org}{www.nextstrain.org}.
This approach is similar in concept to Nextflu \citep{neher2015nextflu} however extended and generalised to different viral pathogens.
There is a growing need for surveillance of non-influenza viruses \citep{tang2017global}, and nextstrain is able to be extended to nearly all viral outbreaks with readily accessible genomic data.

@@

\subsection*{Joint Temporal \& Spatial Visualisation}
Conveying understanding of pathogen evolution through space and time involves filtering large amounts of data into forms that can be easily reasoned with.
Auspice employs linked views into different facets of the data, whereby changes in one view are reflected in all others.
These views allow simultaneous interrogation of phylogenetic and spatial relationships, with additional metadadata such as genotype or serotype expressed through colourings (figure \ref{nextstrain}).
This is coupled with a user-interactable time slider to see how the pathogen has evolved and spread over the course of the epidemic.
By animating the temporal dimension, a high level overview of the entire outbreak is quickly gained.
This approach both conveys the geographical spread of disease and the chaotic nature of epidemics.

\subsection*{Transmission Chain Analysis}
Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probably transmission events given the sampled data, together with a marginal likelihood for each event.
Inferred transmissions are displayed as colours on the tree, with opacity used to convey confidence, and each transmission is also drawn on the map.
More?

@@

\section*{Acknowledgements}
We would like to thank...

@@

This work has been supported by the...
OpenSciencePrize to TB and RAN, the ERC through StG-260686 to RAN.

>>>>

Nextstrain consists of three components:
``Fauna'' is a collection of python scripts to maintain a database representative of all available sequences and related metadata, sourced from public repositories such as NCBI, GISAID and ViPR, as well as github repositories and other sources of genomic data. ``Augur'' is the bioinformatics pipeline, performing subsampling, alignment, phylogenetic inference, temporal dating of ancestral nodes and discrete trait reconstruction across the tree, including inference of the most likely transmission events.
This leverages the maximum likelihood phylodynamic analysis TreeTime \citep{sagulenko2017treetime}, allowing a full analysis of the entire Ebola epidemic ($n=000$) in 0 hours.
These scripts separate generic core functionality from a light virus specific layer such that they are easily co-opted to different pathogens.
Analysed data are then immediately available to interrogate through the visualisation software ``Auspice'', available at \href{www.nextstrain.org}{www.nextstrain.org}.
This approach is similar in concept to Nextflu \citep{neher2015nextflu} however extended and generalised to different viral pathogens.
There is a growing need for surveillance of non-influenza viruses \citep{tang2017global}, and nextstrain is able to be extended to nearly all viral outbreaks with readily accessible genomic data.

@@

\subsection*{Joint Temporal \& Spatial Visualisation}
Conveying understanding of pathogen evolution through space and time involves filtering large amounts of data into forms that can be easily reasoned with.
Auspice employs linked views into different facets of the data, whereby changes in one view are reflected in all others.
These views allow simultaneous interrogation of phylogenetic and spatial relationships, with additional metadadata such as genotype or serotype expressed through colourings (figure \ref{nextstrain}).
This is coupled with a user-interactable time slider to see how the pathogen has evolved and spread over the course of the epidemic.
By animating the temporal dimension, a high level overview of the entire outbreak is quickly gained.
This approach both communicates the geographical spread of disease and the chaotic nature of epidemics.

\subsection*{Transmission Chain Analysis}
Maximum likelihood ancestral state reconstruction of discrete traits such as country or region of isolation allows identification of probably transmission events given the sampled data, together with a marginal likelihood for each event.
Inferred transmissions are displayed as colours on the tree and each transmission is also drawn on the map.
It is important to make known the confidence of any such reconstruction, and we do this in two ways: by matching colour saturation to the confidence of that trait, and by displaying all relevant information when one hovers over the corresponding branch or isolate on the tree.

@@

% please add acknowledgment if applicable.
%\section*{Acknowledgements}
%We would like to thank...
%
%\vspace*{-12pt}

@@

This work has been supported by the OpenSciencePrize to TB and RAN and the ERC through StG-260686 to RAN.
